### 2023-03-02 calculus.tex
Updates from Overleaf

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a}\todo{snapl} extended
with numbers and booleans and empty expression holes. The syntax is given in
\cref{fig:calculus-syntax}. There are two expression languages:
\begin{itemize}
  \item The \textbf{unmarked language}, which is our original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The typing rule for marked addition
expressions then mirrors \textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

@@

\subsection{Polymorphism and Judgemental Structure}
\label{sec:calculus-poly}

@@

\section{Integrating Marking with Editing}
\label{sec:calculus-structured-editing}
\subsection{Hazel Implementation}
We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server. In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we took above). Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser. The supplemental material contains a pre-built implementation.

The result of this effort is the first full-scale typed functional language equipped with a language server that solves the gap problem: all of Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing, are available as long as the program sketch is syntactically well-formed.
%

Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
a structure editor that is able to additionally maintain a form

The previous section presented a formal calculus for error localization and recovery. We now discuss
its usage in a structured editing setting. In particular, the calculus has been integrated into the
Hazel programming environment, a live programming environment with support for typed holes.

\subsection{Implementation in Hazel}
\label{sec:calculus-hazel}

\subsection{Reified Hazelnut Action Semantics}
\label{sec:calculus-hazelnut}

The Hazelnut action calculus introduced by \citet{HazelnutPOPL} gives a structure editing scheme
that employs \emph{expression holes}, denoted $\EEHole$, which represent syntactically incomplete
part of the program. Not only are terms always syntactically well-formed, they are always
\emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
bidirectional action calculus---given by the synthetic action judgment, written
$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
$\hhole{\EMV}{}$, which indicate type errors. \todo{example?}

Hazelnut, however, is unable to cope with actions that require \emph{non-local} hole fixes. For
example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
wrap an arbitrary expression $\EMV$ in a lambda abstraction, i.e. construct
$\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
shadow a previous declaration and require the removal of error holes anywhere in body, but Hazelnut
may only perform such fixes for immediately local holes. Such actions remain undefined, i.e.
unperformable.

By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus, given by the singular action
judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
syntax, and the marking procedure yields statically meaningful terms with error marks, now error
holes, inserted at the appropriate positions as a kind of editor decoration. This solves the problem
outlined above; since marking operates on the entire program after each action, arbitrary
constructions are permitted.

Alternatively, instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut
calculus with scope-based marking. Action operate over \emph{marked expressions}, and error marks
operate as error holes. The system is once again bidirectional, writing (note $\ECMV$ instead of
$\EMV$) the judgments $\ASEAction{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{\AMV}$ and
$\AAEAction{\ctx}{\ECMV}{\ECMV'}{\TMV}{\AMV}$. Then, when wrapping $\ECMV$ into
$\ECLam{x}{\TUnknown}{\ECMV}$, we may simply re-mark the body under the context with $x$ binding
the unknown type:
\[%
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TUnknown}}{\erase{\ECMV}}{\ECMV'}{\TMV'}
  }{
    \ASEConLam{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{x}
  }{ASEConLam}
\]%
This system optimizes over the untyped version by performing marking only when necessary. In
particular, whenever binders are introduced around an expression, marking redraws error holes. The
supplemental material contains a complete formal description of both variants, and the untyped
action semantics and related metatheory are mechanized in Agda.

>>>>

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, and empty expression holes \cite{HazelnutSNAPL}. The syntax is given in
\cref{fig:calculus-syntax}. There are two expression languages:
\begin{itemize}
  \item The \textbf{unmarked language}, which is our original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

The marking rule parallels \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The typing rule for marked addition
expressions then mirrors \textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

@@

\subsection{Polymorphism and Richer Judgemental Structures}
\label{sec:calculus-poly}

@@

\section{Integrating Marking with Editing}
\label{sec:calculus-structured-editing}
\subsection{Hazel Implementation}
\label{sec:calculus-hazel}

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server. In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we developed above). Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser. The supplemental material contains a pre-built implementation.

The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing.
%

Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
a structure editor that is able to additionally offer partial syntactic error recovery.
In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automaticallys insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{tylr}. Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery. In combination with our contributions, this would achieve the Hazel project's stated goal of ensuring that \emph{there are no meaningless editor states}.

\subsection{Fixing Holes in Hazelnut}
\label{sec:calculus-hazelnut}

Earlier versions of Hazel achieved total syntax error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity. The Hazelnut action calculus introduced by \citet{HazelnutPOPL} specifies such a structure editor formally. Notably, this calculus also supports a form of type error localization and recovery: edit actions can automatically insert non-empty holes around type inconsistencies that appear at the cursor after an edit action. This requires specifying a bidirectionally typed action calculus that propagates type and binding information to the cursor, given by the synthetic action judgment, written
$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$.

A significant issue with Hazelnut is that it does not allow (i.e. it leaves undefined) actions that require \emph{non-local} mark (i.e. non-empty hole, in Hazelnut's parlance) insertion or removal. For
example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
wrap an arbitrary expression $\EMV$ in a lambda abstraction, i.e. construct
$\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
shadow a previous declaration and require the removal of error holes in body. The calculus includes manual non-empty hole insertion and removal actions as a workaround.

The fundamental issue is that Hazelnut did not have a total marking system that it could deploy to remark expressions where needed. The marked lambda calculus allows us to solve this problem, in two different ways.

One approach is to define an untyped action calculus, \emph{wholly decoupling} edit actions from typing. Under this simplified design, the action calculus, given by the singular action
judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
syntax, and the total marking procedure yields statically meaningful terms with error marks inserted at the appropriate positions as a kind of editor decoration. This solves the problem
outlined above; since marking operates on the entire program after each action, arbitrary
constructions are permitted.

Alternatively, instead of a wholly untyped action calculus, one may directly integrate re-marking logic into the typed Hazelnut
calculus, making use of the type and scoping information being propagated to re-mark only in scopes where it might be necessary. For example, we can wrap $\ECMV$ into
$\ECLam{x}{\TUnknown}{\ECMV}$ by re-mark the body under the context with $x$ binding
the unknown type:
\[%
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TUnknown}}{\erase{\ECMV}}{\ECMV'}{\TMV'}
  }{
    \ASEConLam{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{x}
  }{ASEConLam}
\]%
The
supplemental material contains a complete formal description of both variants, and the untyped
action semantics and related metatheory are mechanized in Agda. Because of the simplicity of the untyped version, and because marking is defined separately, this Agda mechanization can serve universally as a baseline for correctness of approaches, like our initial design that aim to make use of type and scope information, and perhaps other analyses, to minimize re-marking. We leave a complete assessment of this and other remarking optimization approaches to future work.

### 2023-03-02 let.tex
Updates from Overleaf

To ease the use of products, many languages feature destructuring bindings. In a typed setting, we may want to granularly add type annotations in patterns as well. As it turns out this combination of features is surprisingly tricky to get right! Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

Though the distinction is benign in this example, there is a subtle semantic distinction: synthetic
mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
constraint. This manifests when expressions may have internal type inconsistencies,  as in the
following:
\[%
  \ELet{\PVar{a}}{\EIf{\ETrue}{1}{\EFalse}}{\EMV}
\]%
If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
with an inconsistent branches error following our development of marking above. If instead it is in
analytic position against the unknown type, no mark will be produced.

>>>>

To ease the use of products, many languages feature destructuring bindings. In a typed setting, we may want to granularly add type annotations in patterns as well. As it turns out this combination of features is surprisingly tricky to get right! Let us add let
expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

Though it is benign in this example, there is a subtle semantic distinction: synthetic
mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
constraint. This manifests when expressions may have internal type inconsistencies,  as in the
following:
\[%
  \ELet{\PVar{a}}{\EIf{\ETrue}{1}{\EFalse}}{\EMV}
\]%
If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
with an inconsistent branches error following our development of marking above. If instead it is in
analytic position against the unknown type, no mark will be produced.
